Developing an automated iterative near-term forecasting system for an ecological

### 2018-07-04 manuscript/response_to_review.Rmd

## Associate Editor

**The main issue regards the motivation (reviewer 1) and target audience (reviewer 2) of the application. If this is meant for researchers that are proficient at modeling and coding, it should be made clearer how they benefit from this automated application (wouldn't they already have similar protocols and pipelines in place in their routine research?). If instead, this is for empiricists, or environmental managers, with little coding experience, the manuscript is too technical and the application perhaps should be made more 'user-friendly' (reviewer 1). My suspicion (and reviewer 2's) is that it aims for that increasingly common middle person (the computationally literate ecologist). However, as Reviewer 2 points out, it is not clear how such audience should interact with the defined pipeline and how, for example, they could propose new models. A fourth (very attractive) possibility is that this is meant as an interface for people with different skills and motivations to collaborate (e.g. to foster the feedback between scientific development and applied use). If this is the case, the protocols of interaction need to be clarified, as well as the audience for this article (e.g. is it the computational ecologist who will read this article and propose applied scientists to work with her/him through this system?). Defining the audience clearly should guide how to improve the application and manuscript in a way that it ensures its use and long-term survival as an important short-term forecasting tool.**

We have expanded the last paragraph of the Introduction to address the question of target audience and split the opening paragraph of the Introduction into two paragraphs to provide expanded treatment of the motivation for this approach including the addition of three examples.

The AE and reviewer are correct that we have targeted this paper primarily at the "increasingly common middle person" and "researchers that are proficient at modeling and coding". Thanks to the expanded efforts at computational training, we believe that many researchers have the necessary skills to implement these sorts of approaches, either on their own or as part of a team. And we think that the association between "empiricists" and "little coding experience" is increasingly out of date as demonstrated by the fact that this system was developed in large part by the empiricists collecting the data.

@@

There is definitely a need and demand for short-term forecasting. This was laid out in detail in Dietze et al. (2018; PNAS), but we agree that the current paper benefits from providing more direct treatment of this. We have significantly expanded the opening paragraphs of the introduction to provide specific examples and further emphasize the value of this approach.

**I might also suggest differentiating parameter estimation and model development. Forecasts can improve by adding new models, or by adding new data which then improves parameter estimation in the old models.  The automatic pipeline is great at the second part of this, continually adding data and re-fitting the old models to the new data and then producing new forecasts.  However, since there was so much data in the Portal project to begin with, I doubt that parameter estimates in the models have changed much since the automatic pipeline was established (maybe they have?).  However, in a new project that was just getting started (e.g. NEON), the main improvements in forecast skill early on would be in adding more data.  Eventually, however, the models might stop improving and reach some prediction limit imposed by the structure of the model.  As the authors point out in the discussion, finding this point will be key for improving ecological theory. I think a lot of ecologists think their models are useful, they just need better parameter estimates.  My sense is that actually, most ecological models are just crap for forecasting, it’s not for lack of data! Maybe this is an obvious point, but the role of improved parameter estimation in this whole endeavor kind of gets lost for me.  Does the pipeline even save the fitted model parameters? Would that be worthwhile?**

There are a number of interesting points in this paragraph that we feel are beyond the scope of this current paper. We have already described the importance of new models and how we have facilitated adding them. We have now further expanded this section to describe how those external to the group can contribute improved models. Introducing and discussing the different implications of different forms of error and how they are associated with increased data vs. improvements in models is an interesting subject for forecasting but is out of scope of this paper. It would definitely be something interesting to explore in future work.

@@

**In Figure 1, I feel that something is missing here about how other scientists are supposed to interact with the data and forecasts.  The manuscript makes it pretty clear that this is not just for internal use, but is it actually inviting other labs to add models and make predictions? How specifically would another lab contribute a model to the project? Would they fork the project on github, develop a new model and then do a pull request? Or would they create a separate prediction archive themselves? Maybe add an arrow in Fig 1. showing how new models can be plugged into the forecasting work flow.  Box 1 addresses some of this but I guess I want more details.**

We have added a description of how others can contribute to the Modeling and Forecasting section of the paper. We tried to also add this to the figure, but could not find a way to do so without making the figure overly crowded and complicated.

@@

**Some other miscellaneous questions:**
**Can models be dropped from the ensemble?**

This is not currently done at the moment but is on the development roadmap.

**What if old data needs to be corrected or updated? That could change everything, including the interpretation of the archived predictions. Is there a way to handle that?**

This is a great question about a tricky problem in forecast archiving that we have encountered and are still trying to determine the optimal way to address. One can update the previous forecast files and use the versioning to maintain a copy of the forecasts that contained the underlying error or leave them as they are and store supplementary data indicating that forecasts prior to the selection should not be used in later analyses. Since we aren't happy with any of the current solutions, and this is a fairly detailed point related to archiving, we think it is best to leave this for later discussion.

@@

**Lines 63 to 83:  Could this section include some examples that demonstrate the value of iterative near-term forecasting from other fields? Could it be argued that iterative near term forecasting has led to progress in other fields? Say meteorology and political science?**

We have now mentioned the established benefits for iterative near term forecasting in meteorology and added several citations for readers interested in exploring this further.

**Figure 1: Another point on figure one.  What about model improvement? It seems like there should be a feedback loop somewhere showing how models get refined, added, dropped etc.**

We tried to add a visual representation of this to Figure 1 but could not make it work without making the figure overly crowded. We have added some additional discussion of this to the "Modeling and Forecasting" section of the paper.

@@

**comment on the website: http://portal.naturecast.org/index.html  – Figures need more information/captions.  Which model is producing the blue lines (the forecasts).  Is this the ensemble model?**

Yes. We have clarified this on the website.

**Line 163:  This section here about dependencies seems a little out of place, or at least the transition from the CI section to the details of dependencies management seems abrupt.**

There are two main components to Continuous Analysis as defined by Beaulieu-Jones & Greene (2017): continuous integration and dependency management. We have one paragraph on each in this section on Continous Analysis. We're not sure how else to handle this, but if the reviewer has any specific suggestions we're happy to consider them.

### 2018-07-04 manuscript/portal_iterative_forecasting_ms.Rmd
More minor edits to the revision
* Add citations to living data paper
* Move Boettiger citation to link with Containers not Continuous Analysis
* Small editorial changes
* Add acknowledgement for Heather Bradley

Iterative forecasting is most effective with frequently collected data, since it provides more opportunities for updating model results and assessing (and potentially improving) model performance [Box 1; @dietze2018]. The Portal Project is a long-term ecological study situated in the Chihuahuan Desert (2 km north and 6.5 km east of Portal, Arizona, US). Researchers have been continuously collecting data at the site since 1977, including data on the abundance of rodent and plant species (monthly and twice yearly, respectively) and climatic factors such as air temperature and precipitation (daily) [@brown1998; @ernest2009; @ernest2016]. The site consists of 24 50m x 50m experimental plots. Each plot contains 49 permanently marked trapping stations laid out in a 7 x 7 grid, and all plots are trapped with Sherman live traps for one night each month. For all rodents caught during a trapping session, information on species identity, size, and reproductive condition is collected, and new individuals are given identification tags. This information on rodent populations is high-frequency, uses consistent trapping methodology, and has an extended time-series (475 monthly samples and counting), making this study an ideal case for near-term iterative forecasting.

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@boettiger2015; @beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when the `tscount` package [@liboschik2015], used by two of our forecasting models, was temporarily removed from CRAN (the R package repository) and could not be installed in the usual way. This broke our forecasting pipeline, as we could no longer run models that used that package. To make our pipeline robust to changes in external software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything needed to run a given piece of software, including the operating system. Once created, a software container is basically a time capsule, containing all the software dependencies in the exact state used to develop and run the software. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in the container. We use an existing platform, Docker [@merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to control transitions to new package versions, implementing them only after we have tested them and made any necessary changes to the data processing and analysis code. We use a container created by the Rocker project, which is a Docker image with many important R packages [i.e., the tidyverse packages; @tidyverse] pre-installed  [@boettiger2017]. We add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

Iterative forecasting benefits from frequently updated data so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Both frequent data collection and rapid processing are important for providing timely forecasts. Since we collect data monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database (@yenni2018; Figure 1).

New data are double-entered into Microsoft Excel using the "data validation" feature. The two versions are then compared using an R script to control for errors in data entry. Quality control (QC) checks using the `testthat` R package [@wickham2011] are run on the data to test for validity and consistency both within the new data and between the new and archived data. The local use of the QC scripts to flag problematic data greatly reduces the time spent error-checking and ensures that the quality of data is consistent. The cleaned data are then uploaded to the GitHub-based PortalData repository (https://github.com/weecology/PortalData). GitHub (https://github.com/) is a software development tool for managing computer code development, but we have also found it useful for data management. On GitHub, changes to data can be tracked through the Git version control system which logs all changes made to any files in the repository, giving us a record of exactly of when specific lines of data were changed or added. All updates to data are processed through "pull requests," which are notifications that someone has a modified version of the data to contribute. QA/QC checks are automatically run on the submitted data using continuous integration to ensure that no avoidable errors reach the official version of the dataset.

We also automated the updating of supplementary data tables, including information on weather and trapping history, that were previously updated manually. As soon as new field data is merged into the repository, continuous integration updates all supplementary files. Weather data is automatically fetched from our cellular-connected weather station, cleaned, and appended to the weather data table. Supplementary data tables related to trapping history are updated based on the data added to the main data tables. Using CI for this ensures that all supplementary data tables are always up-to-date with the core data.

@@

The Portal Project has a long history of making its data publicly available so that anyone can use it for forecasting or other projects. Historically, the publication of the data was conducted through data papers [@ernest2009, @ernest2016], the most common approach in ecology; this approach, however, caused years of data latency. With the recent switch to posting data directly to a public GitHub repository (Figure 1) with a CC0 waiver (i.e. no restrictions on data use; https://creativecommons.org/publicdomain/zero/1.0/), data latency for everyone has been reduced to less than one month, making meaningful iterative near-term forecasting possible for not only our group but other interested parties, as well.

@@

Because of the breadth of expertise needed to set up our forecasting pipeline, our effort required a team with diverse skills and perspectives, ranging from software development to field site expertise. It is rare to find such breadth within a single individual, and our system was developed as a collaboration between the lab collecting and managing the data and a computational ecology lab. When teams have a breadth of expertise, communication can be challenging [@winowiecki2011]. We found a shared base of knowledge related to both the field research and computational skills was important for the success of the group. The two labs are part of a joint interdisciplinary ecology group that has a mission of breaking down barriers between field and computational/theoretical ecologists (http://weecology.org). Everyone on the team had received training in fundamental data management and computing skills through a combination of university courses, Software and Data Carpentry workshops [@teal2015], and lab training efforts. In addition, everyone was broadly familiar with the study site and methods of data collection, and most team members had participated in field work at the site on multiple occasions. This provided a shared set of knowledge and vocabulary that actively facilitated interdisciplinary interactions. All members of the team actively participated in the development of the forecasting pipeline. Given the current state of tools for forecasting, forecasting teams require some experience in working with continuous integration and APIs. This means either interdisciplinary teams or additional training will often be required for creating these pipelines until tool development improves. To improve the success of these diverse groups, we believe efforts at providing 'team science' training to scientists interested in forecasting will be beneficial for the success of iterative forecasting attempts for the foreseeable future [@read2016].

@@

We thank Henry Senyondo for help with continuous integration and Hao Ye for discussions and feedback on the manuscript. We thank all of the graduate students, postdocs, and volunteers who have collected the Portal Project over the last 40 years and the developers of the software and tools that made this project possible. This research was supported by the National Science Foundation through grant 1622425 to S.K.M. Ernest and by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through grant GBMF4563 to E.P. White.

@@

The data used in this study is from the Portal Project and is openly available (CC0) on GitHub (https://github.com/weecology/PortalData). Code for reproducing all analyses is available on GitHub (https://github.com/weecology/portalPredictions) and archived on Zenodo [@zenodocode]. Forecasts made by this system are all archived to Zenodo [@zenodoforecasts].

>>>>

Iterative forecasting is most effective with frequently collected data, since it provides more opportunities for updating model results and assessing (and potentially improving) model performance [Box 1; @dietze2018]. The Portal Project is a long-term ecological study situated in the Chihuahuan Desert (2 km north and 6.5 km east of Portal, Arizona, US). Researchers have been continuously collecting data at the site since 1977, including data on the abundance of rodent and plant species (monthly and twice yearly, respectively) and climatic factors such as air temperature and precipitation (daily) [@brown1998; @ernest2009; @ernest2016; @ernest2018]. The site consists of 24 50m x 50m experimental plots. Each plot contains 49 permanently marked trapping stations laid out in a 7 x 7 grid, and all plots are trapped with Sherman live traps for one night each month. For all rodents caught during a trapping session, information on species identity, size, and reproductive condition is collected, and new individuals are given identification tags. This information on rodent populations is high-frequency, uses consistent trapping methodology, and has an extended time-series (475 monthly samples and counting), making this study an ideal case for near-term iterative forecasting.

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when the `tscount` package [@liboschik2015], used by two of our forecasting models, was temporarily removed from CRAN (the R package repository) and could not be installed in the usual way. This broke our forecasting pipeline, as we could no longer run models that used that package. To make our pipeline robust to changes in external software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything needed to run a given piece of software, including the operating system [@boettiger2015]. Once created, a software container is basically a time capsule, containing all the software dependencies in the exact state used to develop and run the software [@boettiger2015]. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in the container. We use an existing platform, Docker [@merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to control transitions to new package versions, implementing them only after we have tested them and made any necessary changes to the data processing and analysis code. We use a container created by the Rocker project, which is a Docker image with many important R packages [i.e., the tidyverse packages; @tidyverse] pre-installed  [@boettiger2017]. We add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

Iterative forecasting benefits from frequently updated data so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Both frequent data collection and rapid processing are important for providing timely forecasts. Since we collect data monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database [@yenni2018; Figure 1].

New data are double-entered into Microsoft Excel using the "data validation" feature. The two versions are then compared using an R script to control for errors in data entry. Quality control (QC) checks using the `testthat` R package [@wickham2011] are run on the data to test for validity and consistency both within the new data and between the new and archived data. The local use of the QC scripts to flag problematic data greatly reduces the time spent error-checking and ensures that the quality of data is consistent. The cleaned data are then uploaded to the GitHub-based PortalData repository (https://github.com/weecology/PortalData). GitHub (https://github.com/) is a software development tool for managing computer code development, but we have also found it useful for data management. On GitHub, changes to data can be tracked through the Git version control system which logs all changes made to any files in the repository, giving us a record of exactly of when specific lines of data were changed or added. All updates to data are processed through "pull requests," which are notifications that someone has a modified version of the data to contribute. QA/QC checks are automatically run on the submitted data using continuous integration to ensure that no avoidable errors reach the official version of the dataset [@yenni2018].

We also automated the updating of supplementary data tables, including information on weather and trapping history, that were previously updated manually. As soon as new field data is merged into the repository, continuous integration updates all supplementary files. Weather data is automatically fetched from our cellular-connected weather station, cleaned, and appended to the weather data table. Supplementary data tables related to trapping history are updated based on the data added to the main data tables. Using CI for this ensures that all supplementary data tables are always up-to-date with the core data [@yenni2018].

@@

The Portal Project has a long history of making its data publicly available so that anyone can use it for forecasting or other projects. Historically, the publication of the data was conducted through data papers [@ernest2009; @ernest2016], the most common approach in ecology; this approach, however, caused years of data latency. With the recent switch to posting data directly to a public GitHub repository (Figure 1) with a CC0 waiver (i.e. no restrictions on data use; https://creativecommons.org/publicdomain/zero/1.0/), data latency for everyone has been reduced to less than one month, making meaningful iterative near-term forecasting possible for not only our group but other interested parties, as well [@ernest2018; @yenni2018].

@@

Because of the breadth of expertise needed to set up our forecasting pipeline, our effort required a team with diverse skills and perspectives, ranging from software development to field site expertise. It is rare to find such breadth within a single individual, and our system was developed as a collaboration between the lab collecting and managing the data and a computational ecology lab. When teams have a breadth of expertise, communication can be challenging [@winowiecki2011]. We found a shared base of knowledge related to both the field research and computational skills was important for the success of the group. The two labs are part of a joint interdisciplinary ecology group that has a mission of breaking down barriers between field and computational/theoretical ecologists (http://weecology.org). Everyone on the team had received training in fundamental data management and computing skills through a combination of university courses, Software and Data Carpentry workshops [@teal2015], and lab training efforts. In addition, everyone was broadly familiar with the study site and methods of data collection, and most team members had participated in field work at the site on multiple occasions. This provided a shared set of knowledge and vocabulary that actively facilitated interdisciplinary interactions. All members of the team actively participated in the development of the forecasting pipeline. Given the current state of tools for automated iterative forecasting, forecasting teams require some experience in working with continuous integration and APIs. This means either interdisciplinary teams or additional training will often be required for creating these pipelines until tool development improves. To improve the success of these diverse groups, we believe efforts at providing 'team science' training to scientists interested in forecasting will be beneficial for the success of iterative forecasting attempts for the foreseeable future [@read2016].

@@

We thank Henry Senyondo for help with continuous integration and Hao Ye for discussions and feedback on the manuscript. We thank all of the graduate students, postdocs, and volunteers who have collected the Portal Project over the last 40 years and the developers of the software and tools that made this project possible. We thank Heather Bradley for all of her logistical support that made this research possible. This research was supported by the National Science Foundation through grant 1622425 to S.K.M. Ernest and by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through grant GBMF4563 to E.P. White.

@@

The data used in this study is from the Portal Project and is openly available (CC0) on GitHub (https://github.com/weecology/PortalData) and archived on Zenodo (@portaldata). Code for reproducing all analyses is available on GitHub (https://github.com/weecology/portalPredictions) and archived on Zenodo [@zenodocode]. Forecasts made by this system are all archived to Zenodo [@zenodoforecasts].
